Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming

### 2025-06-07 paper/draft.tex
Add intro, update abstract and add related works

@@

\maketitle
\begin{abstract}
The development of automatic disease prediction systems is critically important for improving human health and supporting physicians in accurate diagnosis and treatment. Accurate diagnosis is essential for timely intervention and effective therapy, and it can also help prevent further complications and burdensome medical procedures.
Symbolic artificial intelligence has previously been applied in the medical domain \cite{Alviano_2020}. However, its adoption has been limited due to the complex knowledge bases required for such systems to operate.

The proposed method, which combines Large Language Models (LLMs) with Answer Set Programming (ASP), offers a powerful approach by providing a reliable and efficient means of generating a knowledge base that can then be used for disease prediction.
The novelty of our method lies in the combination of these two paradigms. LLMs are capable of encoding medical literature into ASP code with remarkable accuracy and deep comprehension, while ASP contributes structured reasoning and explainability. This combination enables the development of a robust disease prediction system that integrates the \textbf{accuracy and intelligence of language models} with the \textbf{explainability and efficiency of Answer Set Programming}.
Our method demonstrates promising results on the patient dataset to which it was applied.
\end{abstract}

\section{Related work}

\subsection{Large Language Models}
Large language models (LLMs) \cite{zhao2023survey} are advanced computational systems 
designed to understand and generate human language. 
Developed using cutting-edge machine learning techniques, 
these models can produce text that closely mirrors human linguistic patterns.
Modern architectures such as \textit{GPT-4} \cite{openai2023gpt4}, 
are trained on massive corpora spanning diverse domains, 
enabling them to acquire broad, cross-disciplinary knowledge. 
LLMs use deep learning algorithms \cite{Sarker2021} 
and specifically transformer networks \cite{Dosovitskiy2020} 
to analyze and understand the relationships between words, phrases, and sentences. 
By capturing patterns and structures in the training data, 
they can generate coherent and relevant responses based on given prompts.

The term “large” refers to the billions of parameters 
that allow these models to generate context-aware responses. 
The basic premise of a language model is its ability to predict 
the next word (referred to as a token) based on the text it has observed so far, 
according to the context.

@@

\subsection{Challenges in applying LLMs}

LLMs are not without errors \cite{Raj2023, Ruis2023}. 
They can occasionally produce incorrect or biased responses 
that reflect the biases inherent in the training data. 
There have been efforts to mitigate these issues 
through data preprocessing techniques and fine-tuning \cite{Dodge2021}.

Ongoing research seeks to improve the factual accuracy, safety, and interpretability 
of these models, enabling their deployment in high-stakes environments \cite{Ganguli2022}.

\subsection{Answer Set Programming}

Answer Set Programming (ASP) \cite{Eiter2009} is a declarative problem-solving paradigm 
rooted in logic programming and non-monotonic reasoning \cite{Brewka2011}. 
The formal semantics of stable models and the foundational ASP language 
were first introduced by Gelfond and Lifschitz \cite{Gelfond2000, gel88}.

Programming with this approach is done in a family of languages 
sometimes referred to as \textit{AnsProlog} \cite{Gelfond2002}.

@@

\subsection{ASP for disease diagnosis}
ASP offers a powerful method for representing medical knowledge in a declarative and logical manner. 
Some common ASP methods used in the representation of medical knowledge \cite{Vinarti2019}:

\begin{itemize}
    \item \textbf{Rule-Based Representation}: ASP allows for the representation of medical knowledge as rules in the form of logical statements. These rules define relationships between medical entities such as symptoms, diseases, treatments, and patient data. For example, a rule could state that if a patient exhibits specific symptoms, it indicates the presence or absence of certain diseases.
    \item \textbf{Knowledge Base Construction}: ASP provides a framework for constructing a knowledge base that captures medical knowledge. The knowledge base consists of facts, rules, and constraints and may capture a wide range of medical information, including symptom-disease associations, diagnostic criteria, treatment protocols, and patient-specific data.
\end{itemize}

Earlier methods address complex queries over biomedical ontologies 
and databases by extracting relevant information from structured 
knowledge sources \cite{Erdem2011}. Applied in drug discovery, 
these techniques integrate biological databases with concise, 
interpretable explanations for complex queries related to drug development.

\section{Methodology}

\subsection{Architecture}

@@

The goal is to extract a structured knowledge base from unstructured medical text \cite{ré2014feature}.
\Cref{fig:architecture} illustrates the proposed architecture, which follows the key steps below:

\begin{itemize}
    \item Collect medical literature relevant to the diseases or conditions of interest.
    \item Design a suitable prompt to convert the medical literature into ASP code.
    \item Incorporate the ASP code generated by the LLM into the knowledge base.
    \item Add relevant patient data—such as symptoms and examination results—to the knowledge base.
    \item Run an ASP solver to reason over the final knowledge base and produce the diagnosis or set of possible diagnoses as answer sets.
\end{itemize}

\subsection{Prompt Construction}

@@

\subsection{Construction of a Logical Program}

@@

\subsection{Explainability of Diagnoses}

@@

The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}} \cite{Cabalar_2020} 
allows the presentation of justifications regarding the results given by the solver \textit{Clingo}. 
This is achieved by adding a \src{trace} corresponding to the rules of our logic program. 
roviding a clear explanation of how the system reached specific conclusions can help identify potential 
errors in the creation of the knowledge base and provide the system user with a rationale for the final outcome. 
This is particularly important when the application is used in medical contexts.

>>>>

\newcommand{\sys}{\textsc{Sys}\xspace}

@@

\maketitle
\begin{abstract}
Accurate disease prediction is vital for timely intervention, 
effective treatment, and reducing medical complications. 
While symbolic AI has been applied in healthcare, 
its adoption remains limited due to the complexity 
of required knowledge bases.
This work introduces \sys, a framework that combines Large Language Models (LLMs) 
with Answer Set Programming (ASP) to overcome this barrier. 
LLMs extract medical knowledge from text and translate it into ASP code, 
while ASP enables structured, explainable reasoning. 
This integration yields a robust, interpretable prediction system 
that leverages the strengths of both paradigms. 
This approach shows strong performance on real patient data.
% \sys accurately predicts diseases based on symptoms derived from medical literature.
\end{abstract}

% TODO: Add text between sections and paragraphs
\section{Introduction}

The development of automatic disease prediction systems 
is crucial for enhancing healthcare and supporting medical 
professionals in timely and accurate diagnosis. 
Symbolic artificial intelligence (AI), particularly Answer Set Programming (ASP), 
has been explored in this context due to its ability 
to support structured and explainable reasoning. 
However, the construction of expressive knowledge bases 
remains a significant barrier to practical adoption.

This paper introduces \sys, a framework 
that combines Large Language Models (LLMs) with the 
formal reasoning capabilities of ASP.
\sys extracts and encodes medical knowledge from text into ASP code,
enabling interpretable and accurate diagnosis 
of diseases based on patient symptoms.

\section{Background}

\paragraph{\textbf{Large Language Models}}
Large language models (LLMs) \cite{zhao2023survey} are advanced computational systems 
designed to understand and generate human language. 
Developed using cutting-edge machine learning techniques, 
these models can produce text that closely mirrors human linguistic patterns.
Modern architectures such as \textit{GPT-4} \cite{openai2023gpt4}, 
are trained on massive corpora spanning diverse domains, 
enabling them to acquire broad, cross-disciplinary knowledge. 
LLMs use deep learning algorithms \cite{Sarker2021} 
and specifically transformer networks \cite{Dosovitskiy2020} 
to analyze and understand the relationships between words, phrases, and sentences. 
By capturing patterns and structures in the training data, 
they can generate coherent and relevant responses based on given prompts.
The term “large” refers to the billions of parameters 
that allow these models to generate context-aware responses. 
The basic premise of a language model is its ability to predict 
the next word (referred to as a token) based on the text it has observed so far, 
according to the context.

@@

\paragraph{\textbf{Challenges in applying LLMs}}

LLMs are not without errors \cite{Raj2023, Ruis2023}. 
They can occasionally produce incorrect or biased responses 
that reflect the biases inherent in the training data. 
There have been efforts to mitigate these issues 
through data preprocessing techniques and fine-tuning \cite{Dodge2021}.
Ongoing research seeks to improve the factual accuracy, safety, and interpretability 
of these models, enabling their deployment in high-stakes environments \cite{Ganguli2022}.

\paragraph{\textbf{Answer Set Programming}}

Answer Set Programming (ASP) \cite{Eiter2009} is a declarative problem-solving paradigm 
rooted in logic programming and non-monotonic reasoning \cite{Brewka2011}. 
The formal semantics of stable models and the foundational ASP language 
were first introduced by Gelfond and Lifschitz \cite{Gelfond2000, gel88}.
Programming with this approach is done in a family of languages 
sometimes referred to as \textit{AnsProlog} \cite{Gelfond2002}.

@@

\section{Related work}
% we need to add something about LLM + ASP in disease diagnosis
\paragraph{\textbf{ASP for disease diagnosis}}
Answer Set Programming (ASP) has been used to represent 
and reason over medical knowledge in a declarative, logic-based manner. 
Early work applied ASP to biomedical ontologies and structured databases 
to answer complex diagnostic and pharmacological queries with interpretable 
logical justifications \cite{Erdem2011}. In the clinical domain, 
ASP is especially effective for encoding medical rules and patient-specific data, 
enabling precise and explainable decision-making.
Common strategies in ASP-based medical reasoning include:

\begin{itemize}
    \item \textbf{Rule-Based Representation}: ASP allows for the representation of medical 
    knowledge as rules in the form of logical statements. 
    These rules define relationships between medical entities such as symptoms, 
    diseases, treatments, and patient data. For example, a rule could state that 
    if a patient exhibits specific symptoms, it indicates the presence or absence 
    of certain diseases.
    \item \textbf{Knowledge Base Construction}: ASP facilitates the construction of 
    structured knowledge bases composed of facts, rules, and constraints. These can 
    model diagnostic criteria, symptom-disease associations, and therapeutic protocols 
    in a consistent and extensible manner.
\end{itemize}

\paragraph{\textbf{Integrating LLMs with ASP}}

Recent work explores the synergy between LLMs and ASP 
for enhanced diagnostic reasoning. 
These hybrid approaches aim to combine the language understanding of LLMs 
with the precision and explainability of symbolic reasoning. 
A YAML-based interface has been proposed to guide LLMs 
in extracting structured facts from natural language, 
which are then processed by ASP for logical inference \cite{alviano2024llm2asp}. 
Other work uses ASP to validate and trace misleading or inconsistent outputs 
from LLMs in medical contexts \cite{Nguyen2025}. 
A specialized lightweight LLM has also been trained to generate valid ASP code, 
improving the automation of rule-based medical reasoning \cite{coppolillo2024llasp}.
These systems assist in building reliable and clear reasoning pipelines 
for disease diagnosis by combining language understanding with rule-based validation.
\section{Methodology}

\paragraph{\textbf{Architecture}}

@@

The goal is to extract a structured knowledge base from unstructured medical
text \cite{ré2014feature}. \Cref{fig:architecture} illustrates the proposed 
architecture, which proceeds as follows. 
First, medical literature relevant to the diseases or conditions of interest 
is collected. Then, a carefully designed prompt is used to convert this literature 
into ASP code via a large language model. 
The generated ASP code is incorporated into a growing knowledge base, 
which is further enriched with patient-specific data such as symptoms 
and examination results. Finally, an ASP solver reasons over the complete 
knowledge base to produce a diagnosis--or a set of possible diagnoses--in 
the form of answer sets.

\paragraph{\textbf{Prompt Construction}}

@@

\paragraph{\textbf{Construction of a Logical Program}}

@@

\paragraph{\textbf{Diagnostic Explainability}}

@@

The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}} \cite{Cabalar_2020} 
allows the presentation of justifications regarding the results given by the solver \textit{Clingo}. 
This is achieved by adding a \src{trace} corresponding to the logic program's rules. 
roviding a clear explanation of how the system reached specific conclusions can help identify potential 
errors in the creation of the knowledge base and provide the system user with a rationale for the final outcome. 
This is particularly important when the application is used in medical contexts.
